Ecological Limits to Forecasting: Estimating Uncertainty in Critical Transitions with Deep Learning

### 2022-05-16 manuscript/manuscript.Rmd
:pen:

@@

Even the MCMC case illustrates some of the hard limits to ecological forecasting of critical transitions.
Our MCMC forecast assumes the _true_ model structure, the data-generating process, is known, and the forecaster need only infer the posterior distribution of model parameters.
This is a much stronger assumption than that made by the ML models, though it can potentially be justfied on the basis of a mechanistic understanding of the processes involved.
It is important not to confound the MCMC example here with the use of MCMC in process based models of real systems.
In the real world, this is never the case: all models are at best approximations of the underlying processes [@Oreskes1994].
Despite this advantage, even the MCMC forecasts differ from the distribution of the true process.
Because the available data come from only a small region of the dynamical state space, they are consistent with many possible parameterizations of the same model structure -- which creates likelihood ridges and non-identifiability of specific parameter values.
Using more simplified versions of the dynamical processes in question, such as the canonical form of a bifurcation, can mitigate this issue in some cases.
Even when such non-identifiability issues cannot be avoided entirely, they can usually be diagnosed by examining the degree of mixing in MCMC sampling and comparing posterior to prior distributions.
<!-- While similar convergence issues may arise in the estimation of ML models, they may be harder to diagnose. -->

@@

Furthermore, our analysis has focused on the task of making a single forecast prior to the occurrence of a critical transition.
Forecasting is ideally a more iterative process of data assimilation, where forecasts are updated with respect to additional observations, rather than projecting 100s of time steps into the future [@Dietz2018]. 
Updating a forecast after a critical transition has already occurred may be of little use in the context of hysteresis, such as under the saddle node or stochastic transition -- recognizing the alternative stable state only after the system is stuck in that basin will often be considered 'too late'.
Assimilation may be more applicable to the Hopf bifurcation, where additional observations of slowly growing oscillations may lead to more accurate forecasts. 
Such models may even accurately predict the homoclinic bifurcation that occurs when the limit cycle grows too large, eventually hitting a saddle point of zero population size for the host species.
We leave these cases to future exploration.

>>>>

The MCMC case illustrates some of the hard limits to ecological forecasting of critical transitions.
Our MCMC forecast assumes the _true_ model structure, the data-generating process, is known, and the forecaster need only infer the posterior distribution of model parameters.
This is a much stronger assumption than that made by the ML models, though it can potentially be justfied on the basis of a mechanistic understanding of the processes involved.
It is important not to confound the MCMC example here with the use of MCMC in process based models of real systems.
In the real world, this is never the case: all models are at best approximations of the underlying processes [@Oreskes1994].
Despite this advantage, even the MCMC forecasts differ from the distribution of the true process.
Because the available data come from only a small region of the dynamical state space, they are consistent with many possible parameterizations of the same model structure -- which creates likelihood ridges and non-identifiability of specific parameter values.
Using more simplified versions of the dynamical processes in question, such as the canonical form of a bifurcation, can mitigate this issue in some cases.
Even when such non-identifiability issues cannot be avoided entirely, they can usually be diagnosed by examining the degree of mixing in MCMC sampling and comparing posterior to prior distributions.
<!-- While similar convergence issues may arise in the estimation of ML models, they may be harder to diagnose. -->

@@

@@

Furthermore, our analysis has focused on the task of making a single forecast prior to the occurrence of a critical transition.
Forecasting is ideally a more iterative process of data assimilation, where forecasts are updated with respect to additional observations, rather than projecting 100s of time steps into the future [@Dietz2018]. 
Updating a forecast after a critical transition has already occurred may be of little use in the context of hysteresis, such as under the saddle node or stochastic transition -- recognizing the alternative stable state only after the system is stuck in that basin will often be considered 'too late'.
Assimilation may be more applicable to the Hopf bifurcation, where additional observations of slowly growing oscillations may lead to more accurate forecasts. 
Such models may even accurately predict the homoclinic bifurcation that occurs when the limit cycle grows too large, eventually hitting a saddle point of zero population size for the host species.
We leave these cases to future exploration rather than attempting to explore all such variations in a single narrative. 
We have focused on these illustrative examples accompanied by an extensive appendix of reproducible code implemented using well documented and intuitive frameworks.
We encourage the reader to use these appendices as an entry point into further exploration.
